Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-09-07 paper/review/response-to-review.Rmd

The submission "Expanding tidy data principles to facilitate missing data
exploration, visualization and assessment of imputations" presents the R 
package naniar which provides data structures, summaries and visualizations 
for missing data.

Additional comments:

Please add some discussion on what kind of functionality is available in 
other statistical software packages than R in order to better embed the 
contribution not only within R, but the statistical software ecosystem in 
general.

> We have added the following at the end of section 3:

Missing data can be explored and visualised in other software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}. They provide basic table frequencies of missing data, and basic visualisations on missingness patterns. However, none of these software provide functions to specifically visualise missing values in bivariate or multivariate settings, and it is only possible to visualise missing values after performing data analysis, counting and summarising missingness. Imputation methods from mean to regression, EM, and multiple imputation are all available, with out of the box diagnostics provided to assist comparing imputations. \proglang{python} also provides exploratory data analysis for missing data within \pkg{missingno} [@Bilogur2018], and the \pkg{scikit-learn} package provides imputation methods [@scikit-learn].

@@

Please provide a plain R script which when run reproduces all results 
presented in the manuscript. In general purling the Rmd file and editing 
the resulting R file  slightly in order to guide the user what each block 
does, removing knitr specific content,   etc. is suitable.

> We have added a file, "tidy-missingdata-paper.R" that contains the R code from the paper

* The JSS style (https://www.jstatsoft.org/pages/view/style) should be 
followed more closely, especially using the title vs. sentence case, 
JSS-specific markup (\proglang, \pkg), and improving the BibTeX file. Some 
aspects of the style are handled incorrectly by "rticles", though (see
https://github.com/rstudio/rticles/issues/254).

> I have updated to the latest rticles package and checked that the JSS-specific markup has been fixed in the latest submission.

@@

Manuscript style comments:

o For the code layout in R publications, we typically distinguish input/output using Sinput/Soutput (or equivalently CodeInput/CodeOutput). Unless there are special reasons to format it differently, the input should use the text width (up to 76 or 77 characters) and be indented by two spaces, e.g.,

begin{Sinput}
R> example_model <- lm(response ~ variable1 + variable2 + variable3, 
+    weights = w, data = mydata)
\end{Sinput}

> I believe this has been resolved with the latest update to rticles

### 2020-09-07 paper/tidy-missing-data-paper.Rmd
finish draft response to reviewers

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
---

@@

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors. Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focuses on R [@rcore].

@@

Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the "tidyverse" [@tidyverse]. These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto]. This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains. For example, the \pkg{tidytext} [@tidytext] package for text analysis, the \pkg{tsibble} [@wang2020tsibble] package for time series data, and \pkg{tidycensus} [@tidycensus] for working with US census and boundary data.

@@

\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`. So `Var1` has a sibling column, `Var1_imp`, with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from VIM, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.

@@

The MANET (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The MANET software is no longer maintained and cannot be installed. The approach of [@Swayne1998] in the software XGobi, further developed in ggobi [@Cook2007], focussed on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

MissingDataGUI provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually. Using a GUI to explore missing data facilitates rapid insight into missingness structures. However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

VIM (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. However, these visualizations do not map variables to graphical aesthetics, creating friction when moving through analysis workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further exploration.

@@

@@

A data structure facilitating exploration of missing data needs to be simple to reason with and to transport with a data analysis, otherwise it will not be used. A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively. This matrix was used to explore missing values in the interactive graphics library XGobi, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness. The shadow matrix could be interactively linked to the data. However, there are some limitations to the shadow matrix. Namely, the values 0 and 1 can be confusing representations of missing values, since it is not clear if 0 indicates an absence of observation, or the presence of a missing value.

@@

2.  **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>". For example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).

3.  **Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)). It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.

@@

*Nabular* data binds the shadow matrix column-wise to the original data. It is a portmanteau of `NA` and `tabular`. *Nabular* data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)). *Nabular* data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of `var`, as `var_NA` during analysis. *Nabular* data is a snapshot of the missingness of the data. This means when *nabular* data are imputed, those imputed values can easily be identified in analysis (`var` vs `var_NA`). *Nabular* data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

Although a binary missingness matrix, $B$, could be generated during analysis instead of using a *nabular* data structure, there are key advantages to using *nabular* data. Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place. Imputing values on *nabular* data automatically tracks these imputations.

Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2017] and [@Broman2017]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).

Special missing values are not a new idea, and have been implemented in other statistical programming languages, SPSS, SAS, and STATA. These typically represent missing values as a full-stop, `.`, and record special missing values as `.a` - `.z`. These special values from these languages break the tidy principle of one column having one type of value, as they record both the value, and the multivariate missingness state.

@@

Common missing data operations can be considered verbs, in the tidyverse sense. For missing data, these include: **scan**, **replace**, **add**, **shadow**, **impute**, **track**, and **flag**. Data can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with `NA`. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. The data can be augmented with the **shadow matrix** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

@@

This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", `-99`. This is implemented in the function `miss_scan_count()`, which returns a table of occurrences of that value for each variable. A list of common NA values for numbers and characters, can be provided to help check for typical representations of missings. These are implemented in `naniar` as `common_na_numbers` and `common_na_strings`.

@@

Once possible missing values have been identified, these values can be replaced. For example, a dataset could have the values `-99` meaning a missing value. `naniar` implements replacement with the function, `replace_with_na()`. Values -99 could be replaced in the `x` column with: `replace_with_na(dat_ms, replace = list(x = -99))`. For operating on multiple variables, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` makes a conditional change on columns **if** they meet some condition (such as `is.numeric` or `is.character`).

@@

Understanding the missingness structure can be improved by adding summary information alongside the data. For example, in [@Tierney2015], the proportion of missings in a row is used as the outcome in a model to identify variables important in predicting missingness structures. `naniar` implements a series of functions to add these missingness summaries to the data, starting with `add_`. These are inspired by the `add_count` function in `dplyr`, which adds count information for specified groups or conditions. `naniar` provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values, and readable labels for any missing values with `add_label_shadow()` and `add_label_missings()`.

@@

*Nabular* data has the shadow matrix column-bound to existing data. This facilitates visualization and summaries, and allows for imputed values to be tracked. *Nabular* data can be created with `nabular()`:

@@

Unusual or spurious data values are often identified and `flagged`. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. `naniar` provides tools to encode these special types of missingness in the shadow matrix of *nabular* data, using `recode_shadow()`. This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, -99 could be recoded to indicate a broken machine sensor for the variable `x` with the following:

@@

`naniar` does not reinvent the wheel for imputation, instead working with existing methods. However, `naniar` provides a few imputation methods to facilitate exploration and visualization: `impute_below`, `impute_mean`, and `impute_median`. While useful to explore structure in missingness, they are not recommended for use in analysis. `impute_below` imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.

Similar to `simputation`, each `impute_` function returns the data with values imputed. However, `naniar` does not use a formula syntax, instead each function has "scoped variants" `_all`, `_at` and `_if` as in \@ref(verbs-replace-with). `impute_` functions with no scoped variant, (`impute_mean`), will work on a single vector, but not a data.frame. One challenge with this approach is imputed value locations are not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).

@@

To evaluate imputations they need to be tracked. This is achieved by first using `nabular` data, then imputing, and imputed values can then be referred to by their shadow variable, `_NA` (Figure \@ref(fig:track-impute-example)). The code below shows the track pattern, first using `nabular`, then imputing with `impute_lm`. `label_shadow` then adds a label to facilitate identifying missings:

@@

Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the `color` or `fill` aesthetic in ggplot to `any_missing`, a result of the `add_label_shadow()` function. (Figure \@ref(fig:track-impute-example)). Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows shown below in Table \@ref(tab:impute-summary-out), showing similarities and differences in imputation methods.

@@

Missing values are often ignored when plotting data - which is why the data visualization software, MANET, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996]. However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the nabular data structure.

@@

The first step is to get an overview of the extent of the missingness. Overview visualizations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness. The "airquality" dataset (from base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973. We learn from Figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables. These overview plots are created from the shadow matrix in long form (Figure \@ref(fig:nabularfig)). Numerical statistics can also be reported (Section \@ref(num-sum)).

@@

The shadow matrix can be put into long form, allowing both the variables and cases to be displayed using a heatmap style visualization, with `vis_miss()` from \pkg{visdat} [@visdat] (Figure \@ref(fig:nabularfig)). This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D). Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \@ref(fig:gg-miss-case-var)D). Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation. Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.

@@

The number of times observations are missing together can be visualized using an "upset plot" [@Conway2017]. An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, and scales well with more variables. An upset plot can be constructed from the shadow matrix, as shown in Figure \@ref(fig:airquality-upset) which displays the overlapping counts of missings in the airquality data. The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable. This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables Solar.R and Ozone.

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into *nabular* form, then values are imputed below the range of data using `impute_below_all`. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and `Ozone_NA` is mapped to the fill aesthetic. Nabular data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.

**Univariate split by missingness**. The distribution of a variable can be shown according to the missingness of another variable. The shadow matrix part of nabular is used to handle the faceting, and color mapping. Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, using a faceted histogram (B), and an overlaid density (C). This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values. This type of plot can facilitate exploring missing data distributions. For example, we would expect if data were MCAR, for values to be roughly uniformly missing throughout the histogram or density.

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the R package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
---

@@

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors. Tidy data principles are general, but are comprehensively implemented in the \proglang{R} programming language, so this paper focuses on \proglang{R} [@rcore].

@@

Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the \pkg{tidyverse} [@tidyverse]. These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto]. This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains. For example, the \pkg{tidytext} [@tidytext] package for text analysis, the \pkg{tsibble} [@wang2020tsibble] package for time series data, and \pkg{tidycensus} [@tidycensus] for working with US census and boundary data.

@@

\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix "_imp". So "Var1" has a sibling column, "Var1_imp", with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from VIM, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.

@@

The \pkg{MANET} (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category. \pkg{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The \pkg{MANET} software is no longer maintained and cannot be installed. The approach of [@Swayne1998] in the software \pkg{XGobi}, further developed in \pkg{ggobi} [@Cook2007], focussed on multivariate quantitative data. Missingness is incorporated into plots in \pkg{ggobi} by setting them to be 10% below the minimum value.

\pkg{MissingDataGUI} provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually. Using a GUI to explore missing data facilitates rapid insight into missingness structures. However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

\pkg{VIM} (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. However, these visualizations do not map variables to graphical aesthetics, creating friction when moving through analysis workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further exploration.

@@

## Approaches in other languages

Missing data can be explored and visualised in closed source software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}. These provide basic table frequencies of missing data, and basic visualisations on missingness patterns. However, none of these software provide functions to specifically visualise missing values in bivariate or multivariate settings, and it is only possible to visualise missing values after performing data analysis, counting and summarising missingness. Imputation methods from mean to regression, EM, and multiple imputation are all available, with out of the box diagnostics provided to assist comparing imputations. \proglang{python} also provides exploratory data analysis for missing data within \pkg{missingno} [@Bilogur2018], and the \pkg{scikit-learn} package provides imputation methods [@scikit-learn].

@@

A data structure facilitating exploration of missing data needs to be simple to reason with and to transport with a data analysis, otherwise it will not be used. A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively. This matrix was used to explore missing values in the interactive graphics library \pkg{XGobi}, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness. The shadow matrix could be interactively linked to the data. However, there are some limitations to the shadow matrix. Namely, the values 0 and 1 can be confusing representations of missing values, since it is not clear if 0 indicates an absence of observation, or the presence of a missing value.

@@

2.  **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>". For example: "NA_instrument" uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).

3.  **Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)). It makes a clear distinction with "var_NA" being a random variable of the missingness of a variable, "var". This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.

@@

*Nabular* data binds the shadow matrix column-wise to the original data. It is a portmanteau of "NA" and "tabular". *Nabular* data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)). *Nabular* data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of "var", as "var_NA" during analysis. *Nabular* data is a snapshot of the missingness of the data. This means when *nabular* data are imputed, those imputed values can easily be identified in analysis ("var" vs "var_NA"). *Nabular* data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

Although a binary missingness matrix, $B$, could be generated during analysis instead of using a *nabular* data structure, there are key advantages to using *nabular* data. Firstly, missing value labels "NA" and "!NA" are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place. Imputing values on *nabular* data automatically tracks these imputations.

Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2017] and [@Broman2017]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (\@ref(verbs-recode)).

Special missing values are not a new idea, and have been implemented in other statistical programming languages, \proglang{SPSS}, \proglang{SAS}, and \proglang{STATA}. These typically represent missing values as a full-stop, `.`, and record special missing values as `.a` - `.z`. These special values from these languages break the tidy principle of one column having one type of value, as they record both the value, and the multivariate missingness state.

@@

Common missing data operations can be considered verbs, in the tidyverse sense. For missing data, these include: **scan**, **replace**, **add**, **shadow**, **impute**, **track**, and **flag**. Data can be **scanned** to find possible missings not coded as \code{NA}. These values can then be **replaced** with \code{NA}. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. The data can be augmented with the **shadow matrix** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

@@

This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", "-99". This is implemented in the function \code{miss_scan_count()}, which returns a table of occurrences of that value for each variable. A list of common NA values for numbers and characters, can be provided to help check for typical representations of missings. These are implemented in \pkg{naniar} as objects \code{common_na_numbers} and \code{common_na_strings}.

@@

Once possible missing values have been identified, these values can be replaced. For example, a dataset could have the values "-99" meaning a missing value. \pkg{naniar} implements replacement with the function, \code{replace_with_na()}. Values "-99" could be replaced in the "x" column with: \code{replace_with_na(dat_ms, replace = list(x = -99))}. For operating on multiple variables, there are scoped variants for \code{replace_with_na()}: \code{_all}, \code{_if}, and \code{_at}. This means \code{replace_with_na_all()} operates on **all** columns, \code{replace_with_na_at()} operates **at** specific columns, and \code{replace_with_na_if()} makes a conditional change on columns **if** they meet some condition (such as \code{is.numeric()} or \code{is.character()}).

@@

Understanding the missingness structure can be improved by adding summary information alongside the data. For example, in [@Tierney2015], the proportion of missings in a row is used as the outcome in a model to identify variables important in predicting missingness structures. \pkg{naniar} implements a series of functions to add these missingness summaries to the data, starting with \code{add_}. These are inspired by the \code{add_count()} function in \pkg{dplyr}, which adds count information for specified groups or conditions. \pkg{naniar} provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: \code{add_n_miss()}, \code{add_prop_miss()}, \code{add_miss_cluster()}, and \code{add_any_miss()}, respectively (Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values, and readable labels for any missing values with \code{add_label_shadow()} and \code{add_label_missings()}.

@@

*Nabular* data has the shadow matrix column-bound to existing data. This facilitates visualization and summaries, and allows for imputed values to be tracked. *Nabular* data can be created with \code{nabular()}:

@@

Unusual or spurious data values are often identified and "flagged". For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. \pkg{naniar} provides tools to encode these special types of missingness in the shadow matrix of *nabular* data, using \code{recode_shadow()}. This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, -99 could be recoded to indicate a broken machine sensor for the variable "x" with the following:

@@

\pkg{naniar} does not reinvent the wheel for imputation, instead working with existing methods. However, \pkg{naniar} provides a few imputation methods to facilitate exploration and visualization: \code{impute_below()}, \code{impute_mean()}, and \code{impute_median()}. While useful to explore structure in missingness, they are not recommended for use in analysis. \code{impute_below()} imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.

Similar to \pkg{simputation}, each \code{impute_} function returns the data with values imputed. However, \pkg{naniar} does not use a formula syntax, instead each function has "scoped variants" \code{_all}, \code{_at} and \code{_if} as in \@ref(verbs-replace-with). \code{impute_} functions with no scoped variant, (\code{impute_mean()}), will work on a single vector, but not a data.frame. One challenge with this approach is imputed value locations are not tracked. This issue is resolved with *nabular* data covered in Section \@ref(verbs-track).

@@

To evaluate imputations they need to be tracked. This is achieved by first using *nabular* data, then imputing, and imputed values can then be referred to by their shadow variable, "_NA" (Figure \@ref(fig:track-impute-example)). The code below shows the track pattern, first using \code{nabular()}, then imputing with \code{impute_lm()}. \code{label_shadow()} then adds a label to facilitate identifying missings:

@@

Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the \code{color} or \code{fill} aesthetic in ggplot to \code{any_missing}, a result of the \code{add_label_shadow()} function. (Figure \@ref(fig:track-impute-example)). Imputed values can also be compared to complete case data, grouping by \code{any_missing}, and then summarizing, similar to other \pkg{dplyr} summary workflows shown below in Table \@ref(tab:impute-summary-out), showing similarities and differences in imputation methods.

@@

Missing values are often ignored when plotting data - which is why the data visualization software, \pkg{MANET}, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996]. However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the *nabular* data structure.

@@

The first step is to get an overview of the extent of the missingness. Overview visualizations for variables and cases are provided with \code{gg_miss_var()} and \code{gg_miss_case()} (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness. The "airquality" dataset (from base \proglang{R}), is shown, and contains daily air quality measurements in New York, from May to September, 1973. We learn from Figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables. These overview plots are created from the shadow matrix in long form (Figure \@ref(fig:nabularfig)). Numerical statistics can also be reported (Section \@ref(num-sum)).

@@

The shadow matrix can be put into long form, allowing both the variables and cases to be displayed using a heatmap style visualization, with \code{vis_miss()} from \pkg{visdat} [@visdat] (Figure \@ref(fig:nabularfig)). This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D). Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \@ref(fig:gg-miss-case-var)D). Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is \code{vis_miss()} orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation. Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.

@@

The number of times observations are missing together can be visualized using an "upset plot" [@Conway2017]. An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, and scales well with more variables. An upset plot can be constructed from the shadow matrix, as shown in Figure \@ref(fig:airquality-upset) which displays the overlapping counts of missings in the airquality data. The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable. This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables "Solar.R" and "Ozone".

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into *nabular* form, then values are imputed below the range of data using \code{impute_below_all()}. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and "Ozone_NA" is mapped to the fill aesthetic. *Nabular* data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.

**Univariate split by missingness**. The distribution of a variable can be shown according to the missingness of another variable. The shadow matrix part of *nabular* is used to handle the faceting, and color mapping. Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, using a faceted histogram (B), and an overlaid density (C). This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values. This type of plot can facilitate exploring missing data distributions. For example, we would expect if data were MCAR, for values to be roughly uniformly missing throughout the histogram or density.
